\chapter{Accuracy of Methods in MOPAC} 
\index{MOPAC!accuracy}
\index{Accuracy of methods}
\index{MNDO!accuracy of}
\index{MINDO/3!accuracy of}
\index{AM1!accuracy of}
The following tables illustrate the relative accuracy of three semiempirical
methods: MNDO, AM1, and PM3. PM3 is a re-parameterization of the MNDO method,
in which the AM1 form of the core-core interaction is used. In cases where two
sets of MNDO parameters have been published, the more recent set will be used. 

\section{Protocols used in Determining Accuracy}
The selection of species to be used to demonstrate the accuracy of the various
methods presented some difficulty.   As the designer of the AM1 method, I have
an understandable bias to show that AM1 is better than MNDO.  As the sole
author of PM3, I also have an understandable bias to show that PM3 is better
than AM1.  However, in order to be credible from a reader's point of view,
protocols need to be set up which will convince the reader that these biases 
have not affected the discussions or conclusions in this Chapter.

After some thought, I decided to adopt the following protocols:
\begin{enumerate}
\item All species for which reliable experimental data were available would be
used in generating the raw data for the statistical analyses.
\item Experimental data of low or unknown accuracy was used as reference data
if, and only if, a calculation using that data was done.
\item All results of all data would be reported in un-corrected and un-modified
form.
\item Only data for which references were available would be used.
\item Experimental data included some high-level {\em ab initio} results. 
Where {\em ab initio} results were used, the {\em ab initio} method (basis set
plus post-Hartree Fock corrections, if necessary) was of much higher
reliability than the semiempirical method.  In other words, any differences
between {\em ab initio} and semiempirical results would be due to errors in the
semiempirical method.
\end{enumerate}

The second of these protocols needs a word of explanation.  Consider the case
of a compound whose $\Delta H_f$ is of highly questionable accuracy, say 
10$\pm$30 kcal/mol.  Should this system be used in discussions of accuracy? If
the calculated PM3 $\Delta H_f$ was 11 kcal/mol and the calculated AM1 $\Delta
H_f$ was 40 kcal/mol, then PM3 would obviously look better than AM1.  In that
case, my bias would be to include the compound.  Conversely, if the results
were reversed, then my bias would suggest that clearly the reference datum was
very inaccurate,  and therefore should not be included.  This is obviously
faulty reasoning.  Therefore, in order to prevent bias affecting the selection
of compounds to be used in these discussions, the decision of whether to
include any specific species was made before the calculation was done.

By adopting these protocols, and by extensive use of automatic table
generation, and by an absolute minimum  of human intervention, I hope the
reader will be convinced that bias has been kept to an absolute minimum.

In the Tables, the accuracy of the reference data is not given.  This is a
known fault. However, addition of the accuracy data would be a lot of work,
and  has not yet been done.  

\section{Structure of the Tables}
Because of the size of these Tables, finding any specific system can be quite
time-consuming.  To help locate species, the order of occurrence of species
within a Table follows very specific rules.  These are:

\begin{itemize}
\item The order of occurrence is dictated by the empirical formula only.
\item Elements in the empirical formula are arranged as in 
J. O. Cox, G. Pilcher, ``Thermochemistry of Organic and Organometallic
Compounds'', Academic Press, New York, N.Y., 1970, e.g.\
H $<$ C $<$ O $<$ N $<$ S $<$ F $<$ Cl $<$ Br $<$ I $<$ Li $<$ Na $<$ K $<$ 
Rb $<$ Cs $<$ Be $<$ Mg $<$ Ca $<$ Sr $<$ Ba $<$ B $<$ Al $<$ Ga $<$ In $<$ 
Tl $<$ Si $<$ Ge $<$ Sn $<$ Pb $<$ P $<$ As $<$ Zn $<$ Cd $<$ Hg $<$ Sb $<$
Se $<$ Te $<$ Bi.
\end{itemize}
 
\index{$\Delta H_f$!average errors in}
\index{Heat of Formation!average errors in}
Average $\Delta H_f$ errors for different types of compound are given  in
Table~\ref{avehof}, and summarized by element in Table~\ref{elehof}. Because of
the importance of $\Delta H_f$s, the raw data are presented in
Table~\ref{hoftab}.  Users often want to know where methods do not work well,
so the worst predicted $\Delta H_f$ are presented for each method.  These are
shown in Tables~\ref{wemndo}, \ref{weam1} and \ref{wepm3}.

Summarizing geometry errors proved difficult because of the sheer volume of
data. Because bond-lengths are very important, and because of the relatively
small amount of data on angles and dihedrals, only average bond-length errors
are presented. For bond-types, average errors are given in
Table~\ref{avegeos}.  This data is further summarized by element in
Table~\ref{elegeos}.  All the raw data are given in Table~\ref{geotab}.

Although Ionization Potential (I.P.) and dipole moment data is important,  the
importance is much less than $\Delta H_f$ and geometric data, therefore only
summaries are given here, see Tables~\ref{weimndo}, \ref{weiam1} and
\ref{weipm3} for the worst predicted values of I.P.s,  and Table~\ref{eleips}
for a summary by element.  For dipole moments, the corresponding Tables are
\ref{wedmndo}, \ref{wedam1}, \ref{wedpm3}, and \ref{eledip}.

\section{Limitations of MNDO}\index{MNDO!accuracy}
MNDO is the oldest of the three methods surveyed here, and as a direct result,
is the least accurate. MNDO has many advantages over earlier semiempirical
methods, but as MNDO is such a large improvement over those methods,
enumeration of MNDO's good points would be invidious to the other methods.
Instead, only the limitations likely to be encountered by users will be
mentioned. The principal drawbacks to MNDO are:
\begin{enumerate}
\item Sterically crowded molecules are too unstable.   Example: neopentane.
\item Four-membered rings are too stable.   Example: cubane.
\item The hydrogen bond is virtually non-existent.   Example: water dimer.
\item Hypervalent compounds are too unstable.   Example: sulfuric acid.
\item Activation barriers are generally too high.
\item Non-classical structures are predicted to be unstable relative to the 
classical structure.   Example: ethyl radical.
\item   Oxygenated substituents on aromatic rings are out-of-plane.   Example: 
nitrobenzene.
\item The peroxide bond is systematically too short by about 0.17\AA.
\item The C--O--C angle in ethers is too large by about 9$^\circ$. 
\end{enumerate}

\section{Accuracy of AM1}\index{AM1!accuracy}
AM1 is a distinct improvement over MNDO, in that the overall accuracy is
considerably improved. Specific improvements are:
\begin{enumerate}
\item The strength of the hydrogen bond in the water dimer is 5.5 kcal/mol, in
accordance with experiment.
\item Activation barriers for reaction are markedly better than those of MNDO.
\item Hypervalent phosphorus compounds are considerably improved relative to
MNDO.
\item In general, errors in $\Delta H_f$ obtained using AM1 are about 40\%
less than those given by MNDO.
\end{enumerate}

Unfortunately, with this improvement a few deficiencies were introduced. The
most important of these are:
\begin{enumerate}
\item\index{Phosphorus!AM1 fault} AM1 phosphorus has a spurious and very sharp
potential barrier at 3.0 \AA ngstroms. The effect of this is to distort
otherwise symmetric geometries and to introduce spurious activation barriers. A
vivid example is given by P$_4$O$_6$, in which the notionally equal P--P bonds
are predicted by AM1 to differ by 0.4 \AA . This is by far the most severe
limitation of AM1.
\item Alkyl groups have a systematic error due to the heat of formation of the
CH$_2$ fragment being too negative by about 2 kcal/mol.
\item Nitro compounds, although considerably improved relative to MNDO, are
still systematically too positive in energy.
\item The lowest-energy conformer for ethanol is predicted to have a gauche 
C--C--O--H angle.  High level {\em ab initio} calculations, and experimental 
evidence, suggests that the correct geometry is trans.  This error was 
reported by Andreas Jabs, at Martin-Luther-Universitaet Halle-Wittenberg,
Halle/Saale, Germany.
\item The peroxide bond is still systematically too short by about 0.17\AA.
\end{enumerate}

\section{Accuracy of PM3}\index{PM3!accuracy}
PM3 is a distinct improvement over AM1. Evidence of this is:
\begin{enumerate}
\item Hypervalent compounds are predicted with considerably improved accuracy.
\item Overall errors in $\Delta H_f$ are reduced by about 40\% relative to AM1.
\end{enumerate}

\section{Faults and Errors in PM3}
Although PM3 is more accurate in predicting the $\Delta H_f$ of many systems
than either AM1 or MNDO, it does have some limitations. The more important of
these are:

\begin{enumerate}
\item The barrier to rotation in formamide is practically non-existent. In
part, this can be corrected by the use of the \comp{MMOK} option, see
Section~\ref{mmok}. Fault discovered by Prof Tim Clark.
\item Minima are predicted for which no experimental systems are known.  By
inference, these minima are spurious~\cite{klebe}.
\item Proton affinities are inaccurate.  The distribution of proton affinities
is essentially ``shotgun.''  In other words, there is a high, non-systematic
scatter from the experimental values compared with all other popular
methods~\cite{schmiedekamp} (including AM1).
\item CBr$_4$ is predicted to have a D$_{2d}$ rather than T$_d$ geometry (W.\
Thiel, private comm.).
\item The lowest-energy conformer for ethanol is predicted to have a gauche
C--C--O--H angle.  High level {\em ab initio} calculations, and experimental
evidence, suggests that the correct geometry is trans.  This error was reported
by Andreas Jabs, at Martin-Luther-Universitaet Halle-Wittenberg, Halle/Saale,
Germany.
\item Some polycyclic rings are predicted to have butterfly shapes rather than
being flat.  
\item Almost all $sp^3$ nitrogen systems are predicted to be pyramidal, in
variance with experiment and other methods.
\item The charge on nitrogen atoms are most likely of incorrect sign  and of
unrealistic magnitude. This was due to faults in the parameterization.
\item H-non-bonded H contacts are too short, at about 1.7\AA . They should be
more than 2\AA .  One consequence is that hydrogen bonds are too short by about
0.1\AA .
\item The LUMO of methane is of symmetry $a_1$ instead of the correct symmetry 
of $t_2$. 
\end{enumerate}

The tables given here have been carefully checked for accuracy. In many
instances the calculated results differ from those of earlier publications. In
most cases, the difference has been investigated and found to be due to errors
in the earlier work. This does not, of course, mean that the tables are fully
accurate, but only that care has been taken to minimize error.

% \section{Comparison of Accuracy of MNDO, AM1, and PM3}
\input{mndo_errors}
\input{am1_errors}
\input{pm3_errors}
\clearpage
\input{avehof}
\clearpage
\input{elehof}
\clearpage
\input{avegeos}
\clearpage
\input{elegeos}
\clearpage
\input{mndoi_errors}
\clearpage
\input{am1i_errors}
\clearpage
\input{pm3i_errors}
\clearpage
\input{eleips}
\clearpage
\input{mndod_errors}
\clearpage
\input{am1d_errors}
\clearpage
\input{pm3d_errors}
\input{eledip}
\clearpage

\section{Comparison of Calc'd and Observed $\Delta H_f$ for MNDO, AM1, and PM3}
The following Table lists experimental $\Delta H_f$, in kcal/mol, for  1200
species.  In order to save space, only the difference between the calculated
and observed $\Delta H_f$ is reported.  The calculated $\Delta H_f$ is,
however, easily obtained by adding the difference to the experimental value. 
Thus, the $\Delta H_f$ of  methane, CH$_4$,  calculated using MNDO, is
5.95+(--17.89) = --11.96 kcal/mol.

All the results reported in this and the following Table can be easily
reproduced using MOPAC.  In some cases, care must be taken to ensure that the
correct spin state and symmetry are specified.  For example, triplet oxygen
must be specified using \comp{OPEN(2,2) TRIPLET}, in order for the correct
$\Delta H_f$ to be obtained.   Most open-shell systems will not be indicated as
open-shell.

Results for the iodine atom are missing.  Although MNDO, AM1, and PM3 all give
a zero error, the PM3 result is incorrect.  The correct PM3 result for the
$\Delta H_f$ of iodine atom is -46.87 kcal/mol.  This can be easily checked by
running an iodine atom using keywords \comp{PM3 OPEN(7,4)}.

Many reference data are of highly questionable accuracy, for example the sulfur
halides, and some other reference data refer to highly exotic species, e.g.\
N$^{2+}$. Results for calculations on all these systems are reported, as
required by the protocols for determining accuracy of methods.  If only those
systems for which accurate data were used, and the choice of system limited to
conventional molecules, then the average errors reported in the earlier Tables
would decrease markedly.

\subsection*{Notes on the Table}
\begin{itemize}
\item The $\Delta H_f$ of COI is probably incorrect.  Given that the
experimental $\Delta H_f$s of CO and I are -26.4 and +25.5 kcal/mol,
respectively, and  the $\Delta H_f$ of COI is 63.6 kcal/mol, then the CO-I bond
has a `strength' of -64.5 kcal/mol.  This is unlikely.
\item $\Delta H_f$ for the elements are included. Isolated atoms are used in
the NDDO methods as the standard, and the defined $\Delta H_f$ should be
exactly zero.  For PM3 several elements have negative $\Delta H_f$.  This is a
result of  PM3 incorrectly predicting the atomic configuration.
\end{itemize}

\input{hoftab}

\section{Comp. of Calc'd and Exp'l Geometries for MNDO, AM1, and PM3}

The following Table lists over 1100 experimental bond lengths, angles,  and
dihedrals or torsion angles, and the errors in the prediction of these
quantities for each method.

Reference geometry data are taken from many sources, mainly from X-ray and
microwave structures. No differentiation is made between these various sources,
although microwave structures are preferred to X-ray. Further, no
differentiation is made between various definitions of bond lengths. Thus the
bond length corresponding to the bottom of the harmonic well, $r_o$, is not
distinguished from the equilibrium bond length, or from the bond length
corresponding to the system in which the zero point energy is included. For
semiempirical methods, the average error in bond lengths is large relative to
the differences between the bond lengths obtained using the various
definitions.

\subsection*{Notes on the Table}
\begin{itemize}
\item The AM1 results for P$_4$O$_6$ and P$_4$O$_{10}$ are incorrect. The
experimental geometry was used to start the calculation.  In order to generate
the lowest energy geometry, the systems must first be distorted.
\item For salicylaldoxim, the bonds reported are the internal hydrogen bond and
a non-bonding O-N interaction.
\item Several 180$^{\circ}$ angles are reported. Most of these correspond to
symmetry-defined angles.  In some instances, e.g., MgCl$_2$, the calculated
angle is incorrect.  The protocols for this Chapter require {\em all} such
results to be reported.
\item The PM3 prediction for the bond-length in HgI is in error by over 10\AA .
HgI is predicted, by PM3, to consist of a mercury and an iodine atom.  The
fault can be traced to the incorrect PM3 prediction for the ground state of the
iodine atom. Removal of this one error would lower the average errors in Hg
bond lengths from 0.586\AA\ to 0.054\AA , and the average error in I bond
lengths from 0.371\AA\ to 0.091\AA .  These, in turn, result in the average
bond-length error in PM3 dropping from 0.045\AA .  If that were done, then PM3
would become the most accurate method for the prediction of bond-lengths. 
Again, the protocols involved preclude this being done at this time.
\end{itemize}

\input{geotab}
