\section{Geometry optimization}\index{Geometry!BFGS optimizer}
The default geometry optimizer in MOPAC uses Baker's EigenFollowing method.  If
this is {\em not} wanted, for example, if there is a need to reduce memory
demands, then the Broyden Fletcher Goldfarb Shanno method can be used.

The most common use of MOPAC is for geometry optimization. This involves
starting with an approximation to the desired geometry and, by calculating the
forces acting on the system, changing the geometry so as to lower the total
energy. The objective of geometry optimization is to achieve a structure in
which all the atoms are at equilibrium, that is, one in which the forces acting
on every atom are very small, and in which the second derivatives are
everywhere positive.  Such a geometry is called a ground state stationary
point.

\input{t_ef}
\input{t_bfgs}

\subsection{Optimization of one unknown}
If a system has exactly one coordinate to be optimized, then obviously one
line-search will optimize the geometry.  Because of this, the geometry
optimization is done a little differently.  Given the initial geometry, the
$\Delta H_f$ is calculated, and the line-search started. Unlike the normal
line-search, however, the search is not stopped when the minimum is almost
reached, instead, the minimum is located with quite high precision.  After the
line-search is complete, the gradients are not \index{GRADIENTS} calculated
(unless requested by \comp{GRADIENTS}). Instead, it is assumed that the
gradient is small, and the results are output.  This saves some time. However,
if \comp{GRADIENTS} is {\em not} present, and the geometry is not at a
stationary point (because other coordinates are not optimized), then the
warning message that the geometry is not at a stationary point will not be
printed.

\index{Line search|)}

\input{t_geometry2}
